---
title: "week9 exercise"
author: "Xiang Li"
date: "2023/12/29"
output: pdf_document
---
```{r}
library(rbenchmark)
```
# Exercise 1
```{r}
set.seed(9)
x1 = rbinom(n = 20000, size = 5, prob = 0.3)
x2 = rgamma(n = 30000, shape = 2, rate = 1)
```
## 1
```{r}
mean(x1)
mean(x2)
```
`rbinom` generates ramdom samples that follow the specified binomial distribution with parameter `size` and `prob`.

`rgamma` generates ramdom samples that follow the specified gamma distribution with parameter `shape` and `rate`.

## 2
```{r}
my_mean = function (x) {
  n = length(x)
  sum_x = 0
  for (i in 1:n){
    sum_x = sum_x + x[i]
  }
  mean_x = sum_x/n
  return(mean_x)
}
my_mean(x1)
```
## 3
```{r}
ben1 = benchmark('mean_x1' = {mean(x1)}, 'my_mean_x1' = {my_mean(x1)}, replications = 500)
ben1
ben2 = benchmark('mean_x2' = {mean(x2)}, 'my_mean_x2' = {my_mean(x2)}, replications = 500)
ben2
```
Using `mean()` is faster. The execution time of `my_mean()` is more than 16 times longer than `mean()`.

# Exercise 2
```{r}
set.seed(9)
n = 2000
p = 500
m1 = matrix(rnorm(n*p, mean = 4.7, sd = 0.5), ncol = p)
```
## 1
```{r}
ben = benchmark(
        'apply' = {apply(m1, 2, mean)},
        'colMeans' = {colMeans(m1)},
        'pre_for' = {
          c_n = ncol(m1)
          c_mean_v = rep(NA, c_n)
          for (c in 1:c_n){
            c_mean_v[c] = mean(m1[,c])
          }
        },
        'for' = {
          c_n = ncol(m1)
          c_mean_v = c()
          for (c in 1:c_n){
            c_mean_v[c] = mean(m1[,c])
          }
        },
        replications = 100
)
ben
```
`colMeans()` is fastest and `apply()` is slowest.

# Exercise 3
```{r}
1 - pbinom(44, size = 127, prob = 0.38)
```
# Exercise 4
## 1
```{r}
dpois(7, lambda = 6)
```
## 2
```{r}
0
```
Gamma distribution is a continuous distribution. Thus, for one point, the probability is equal to 0.

## 3
```{r}
ppois(4, lambda = 6) - ppois(2, lambda = 6)
```
## 4
```{r}
pgamma(3, shape = 3, rate = 2) - pgamma(1, shape = 3, rate = 2)
```
## 5
```{r}
ppois(5, lambda = 6)
```
## 6
```{r}
ppois(3, lambda = 6) + pgamma(10, shape = 3, rate = 2)
```
# Exercise 5
## 1
```{r}
set.seed(123)
x = rnorm(1e4, mean = 3, sd = 1.4)
y = rbeta(1e4, shape1 = 2, shape2 = 2)
```
## 2
```{r}
z = x/y
```
## 3
```{r}
mean(z)
var(z)
```
