---
title: "week7 exercise"
author: "Xiang Li"
date: "2023/11/16"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 55), tidy = TRUE)
```
```{r}
library(faraway)
library(MASS)
```
# Exercise 2
```{r}
f = function (df){
  num_col = sapply(df, is.numeric)
  char_col = sapply(df, is.character)
  fac_col = sapply(df, is.factor)
  type = num_col | char_col | fac_col
  if (length(type[type]) == 0){
    return('This dataframe contains no numeric, character or factor values')
  }
  else{
    num_df = df[, num_col]
    if (ncol(num_df) > 0){
      for (i in 1: ncol(num_df)){
      i_name = colnames(num_df)[i]
      hist(num_df[, i], breaks = 100, xlab = i_name, xlim = c(min(num_df[, i]), max(num_df[, i])), col = 'skyblue', main = paste0('Histogram of ', i_name), probability = TRUE)
      }
    }
    char_df = df[, char_col | fac_col]
    if (ncol(char_df) > 0){
      for (i in 1: ncol(char_df)){
      i_name = colnames(char_df)[i]
      count = aggregate(char_df[, 1, drop = FALSE], by = list(name=char_df[, i]), FUN = length)
      colnames(count)[2] = 'count'
      barplot(height = count$count, names.arg = count$name, xlab = i_name, col = 'lightgreen', ylim = c(0, max(count$count)), main = paste0('Barplot of ', i_name))
      }
    }
  }
}
```
```{r}
f(gala)
```
```{r}
f(amlxray)
```

# Exercise 3
## 1
```{r}
mean_v = apply(state.x77, 2, mean)
median_v = apply(state.x77, 2, median)
sd_v = apply(state.x77, 2, sd)
skew_v = apply(state.x77, 2, psych::skew)
sum_sta_df = data.frame(mean=mean_v, median=median_v, standard_deviation=sd_v, skewness=skew_v)
sum_sta_df
```
## 2
```{r}
cns = colnames(state.x77)
for (i in 1:ncol(state.x77)){
  i_name = cns[i]
  hist(state.x77[, i], breaks = 100, xlab = i_name, xlim = c(min(state.x77[, i]), max(state.x77[, i])), col = 'skyblue', main = paste0('Histogram of ', i_name), probability = TRUE, plot = TRUE)
}
```

## 3
The Income, Life Exp and Murder are almost symmetric.Population, Illiteracy and Area are positive skewed. HS Grad and Frost are negative skewed.

## 4
```{r}
plot(x = state.x77[, 'Illiteracy'], y = state.x77[, 'Life Exp'], type = 'p', col = 'orange', xlab = 'Illiteracy', ylab = 'Life Exp', main = 'Illiteracy vs Life Exp')
```
With the increase of Illiteracy rate, the life expectancy decreases.

## 5
```{r}
cor.test(x = state.x77[, 'Illiteracy'], y = state.x77[, 'Life Exp'], alternative = 'two.sided', method = 'pearson')
```
Lower Illiteracy rate means higher education level, which will improve people's life quality and increase life expectancy.

# Exercise 4
```{r}
c1 = c(0, 1)
c2 = expand.grid(c1, c1)
c2s = rowSums(c2)
```
## 1
```{r}
hist(c2s, breaks = 3, main = 'Histogram of c2s')
dim(c2)
```

`expand.grid()` lists all combinations from given vectors. `rowSums()` lists the sum of each combination from given vectors.

## 2
```{r}
c5 = expand.grid(c1, c1, c1, c1, c1)
c5s = rowSums(c5)
hist(c5s, breaks = 6, main = 'Histogram of c5s')
dim(c5)
```
The distribution of c5s gets a peak in center.

## 3
```{r}
c15 = expand.grid(replicate(15, c1, simplify = F))
c15s = rowSums(c15)
hist(c15s, breaks = 16, main = 'Histogram of c15s')
dim(c15) == c(2**15, 15)
```
## 4
```{r}
sta_ls = list(range=range(c15s), mean=mean(c15s), std=sd(c15s))
sta_ls
c(sta_ls$mean, 15*0.5)
c(sta_ls$std, (15*0.5*0.5)**0.5)
```
## 5
```{r}
rn_v = rnorm(1e3, mean = sta_ls$mean, sd = sta_ls$std)
hist(rn_v, breaks = 16, xlim = c(min(rn_v), max(rn_v)), main = 'Histogram of rn_v', probability = TRUE)
```
The distribution of $10^3$ random numbers is similar to the distribution of c15s.

## 6
```{r}
rn_v1 = rnorm(1e3, mean = sta_ls$mean-0.5, sd = sta_ls$std)
hist(rn_v1, breaks = 16, xlim = c(min(rn_v1), max(rn_v1)), main = 'Histogram of rn_v1', probability = TRUE)
curve(expr = dnorm(x, sta_ls$mean-0.5, sta_ls$std), add = TRUE)
```

# Exercise 5
```{r}
fib = c(1, 1)
i = 2
fib_sum = sum(fib)
while (fib_sum < 22000){
  i = i+1
  fib[i] = fib[i-2] + fib[i-1]
  fib_sum = sum(fib)
}
length(fib)
```
