---
title: "week8 exercise"
author: "Xiang Li"
date: "2023/12/28"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 55), tidy = TRUE, error = TRUE)
```
```{r}
library(brolgar)
library(vioplot)
library(mvtnorm)
library(corrplot)
```
# Exercise 1
## 2
```{r}
vc = function(v){
  if (! is.numeric(v)) {
    stop('Function input should be a numeric vector!!!')
  }
  else if (length(v[is.na(v)]) > 0) {
    warning('The supplied vector contains missing values!!!')
  }
  else {
    result = v-mean(v)
    return(result)
  }
}
vs = function(v){
  if (! is.numeric(v)) {
    stop('Function input should be a numeric vector!!!')
  }
  else if (length(v[is.na(v)]) > 0) {
    warning('The supplied vector contains missing values!!!')
  }
  else {
    result = v/sd(v)
    return(result)
  }
}
vn = function(v){
  if (! is.numeric(v)) {
    stop('Function input should be a numeric vector!!!')
  }
  else if (length(v[is.na(v)]) > 0) {
    warning('The supplied vector contains missing values!!!')
  }
  else {
    result = (v-mean(v))/sd(v)
    return(result)
  }
}
```
## 3
```{r}
obj1 = matrix(1:10, 5, 2)
obj2 = c(5, 7, 10, -25)
obj3 = c(42, NA, 3, 7)
obj4 = c(pi, 42, 'apple', sqrt(3))
```
```{r}
vc(obj1)
```
```{r}
vc(obj2)
```
```{r}
vc(obj3)
```
```{r}
vc(obj4)
```
```{r}
vs(obj1)
```
```{r}
vs(obj2)
```
```{r}
vs(obj3)
```
```{r}
vs(obj4)
```
```{r}
vn(obj1)
```
```{r}
vn(obj2)
```
```{r}
vn(obj3)
```
```{r}
vn(obj4)
```
# Exercise 2
## 2
```{r}
sk = function(v){
  v_bar = mean(v)
  result = mean((v - v_bar) ** 3)/(var(v)**(3/2))
  return(result)
}
gen_sk = function(v){
  if (!(is.vector(v) | is.matrix(v))) {
    stop('Wrong type of input!!!')
  }
  else if (any(is.na(v))){
    warning('The supplied vector contains missing values!!!')
  }
  else if (is.matrix(v)){
    warning('Supplied input is a matrix. Skewness by column returned!!')
    return(apply(v, 2, FUN = sk))
  }
  else{
    return(sk(v))
  }
}
gen_sk(matrix(rnorm(100), nrow = 20, byrow = TRUE))
```
# Exercise 3
```{r}

heights = as.data.frame(heights)
wages = as.data.frame(wages)
```
## 1
```{r}
height1980_df = heights[heights$year == 1980, ]
c_n = length(unique(height1980_df$continent))
par(bty = 'l')
boxplot(height_cm ~ continent, data = height1980_df, horizontal = TRUE, main = 'Horizontal Boxplot', col = rainbow(c_n))
par(bty = 'l')
boxplot(height_cm ~ continent, data = height1980_df, horizontal = FALSE, main = 'Vertical Boxplot', col = rainbow(c_n))
```
```{r}
par(bty = 'l')
vioplot(height_cm ~ continent, data = height1980_df, horizontal = TRUE, main = 'Horizontal Vioplot', col = rainbow(c_n))
par(bty = 'l')
vioplot(height_cm ~ continent, data = height1980_df, horizontal = FALSE, main = 'Vertical Vioplot', col = rainbow(c_n))
```

## 2
```{r}
wagesl2_df = wages[wages$xp <= 2, ]
c_n = length(unique(wagesl2_df$high_grade))
par(bty = 'l')
vioplot(ln_wages ~ high_grade, data = wagesl2_df, horizontal = TRUE, main = 'Horizontal Vioplot', col = rainbow(c_n))
```

# Exercise 4
## 1
```{r}
set.seed(8)
Sigma1 = matrix(c(1, 0.5, -0.3, 0.5, 1, -0.6, -0.3, -0.6, 1), 3, 3)
m1 = rmvnorm(500, mean = rep(0, 3), sigma = Sigma1)
Sigma2 = matrix(c(1, 0.4, 0.4, 1), 2, 2)
m2 = rmvnorm(500, mean = rep(0, 2), sigma = Sigma2)
all.vars = cbind(m1, m2)
colnames(all.vars) = paste0('X', 1:5)
```
The code is generating 5 variables which have 500 samples for each one. The first 3 variables follow 3-variate normal distribution with parameter $\mu = [3, 3, 3]$ and $\sigma = Sigma1$. The last 2 variables follow 2-variate normal distribution with parameter $\mu = [2, 2]$ and $\sigma = Sigma2$.

## 2
```{r}
cor_tab = cor(all.vars)
cor_tab
```
## 3
```{r}
cor_tab[1:3, 1:3]
cor_tab[4:5, 4:5]
```
The correlations between $X_1$, $X_2$, $X_3$ are almost equal to Sigma1. The correlations between $X_4$, $X_5$ are almost equal to Sigma2.

## 4
```{r}
corrplot(cor_tab, method="circle")
```

## 5
```{r}
corrplot(cor_tab, method="circle", col = colorRampPalette(c("red", "darkgreen"))( 100))
```
