---
title: "exercise3"
author: "Xiang Li"
date: "2024/2/20"
output: pdf_document
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(tidy_opts = list(width.cutoff = 55), tidy = TRUE)
```
```{r}
## Install and invoke packages
if(!"dagitty" %in% .packages(all = TRUE)) install.packages("dagitty")
if(!"ggplot2" %in% .packages(all = TRUE)) install.packages("ggplot2")
if(!"GGally" %in% .packages(all = TRUE)) install.packages("GGally")
if(!"gridExtra" %in% .packages(all = TRUE)) install.packages("gridExtra")

## Load packages
library(dagitty)
library(ggplot2)
library(GGally)
library(gridExtra)
```
# Exercise 1
```{r}
## Specify the DAG
g.postulated = dagitty('dag {
  C   [pos="-4.000,-3.000"]
  CS  [pos="-4.000,3.000"]
  FL  [pos="1.000,-3.000"]
  I   [outcome,pos="2.500,3.000"]
  IGP [pos="-1.000,3.000"]
  NMF [pos="2.500,-1.000"]
  PGP [pos="-1.000,-2.000"]
  PI  [pos="-4.000,1.000"]
  TM  [pos="-4.000,-1.000"]
  WUE [exposure,pos="-1.000,0.000"]
  C   -> { FL TM }
  CS  -> { IGP PI }
  FL  -> { NMF PGP }
  IGP -> I
  NMF -> { I IGP }
  PGP -> WUE
  TM  -> { IGP PI }
  WUE -> IGP
  }
')

## Plotting DAG
plot(g.postulated)
```

## 1
{CS, TM, WUE, NMF}

## 2
{IGP, CS, TM, C, WUE, PGP, FL, NMF}

## 3
{IGP, I}

## 4
{PI, IGP, I}

## 5
{C->FL->NMF->I, C->FL->NMF->IGP->I, C->FL->PGP->WUE->IGP->I, C->TM->IGP->I}

## 6
Yes.

# Exercise 2
## 7
{}

## 8
{IGP, NMF}

## 9
```{r}
impliedConditionalIndependencies(g.postulated)
```
## 10
```{r}
## Load and inspect data
load("sportsData.RData")
ggpairs(data)+ theme_bw()
```
```{r}
## Perform local tests
r = localTests(g.postulated, data)

## Adjust using Holm-Bonferroni correction
## Retain only those tests with adjusted p-value < .05
r = r[p.adjust(r$p.value) < 0.05, ]

## Plot conditional independencies inconsistent with data
r = r[order(r$p.value), ]
par(mar = c(4, 8.5, 1.5, 0.5))
plotLocalTestResults(r, bty = "n", xlim = c(-0.4, 0.4), ylim = c(0.1, 3.1),
                     axis.pars = list(las = 1, lty = 0))
```

Because it makes no assumption of independence between the consecutive tests.

## 11
Given {C}, TM and WUE are not independent.
Given {FL}, TM and WUE are not independent.
Given {PGP}, TM and WUE are not independent.

## 12
Find a new path between TM and WUE and then find d-seperating sets and check the conditional independencies again.

# Exercise 3
```{r}
## Amended DAG
g.amended = dagitty('dag {
  C   [pos="-4.000,-3.000"]
  CS  [pos="-4.000,3.000"]
  FL  [pos="1.000,-3.000"]
  I   [outcome,pos="2.500,3.000"]
  IGP [pos="-1.000,3.000"]
  NMF [pos="2.500,-1.000"]
  PGP [pos="-1.000,-2.000"]
  PI  [pos="-4.000,1.000"]
  TM  [pos="-4.000,-1.000"]
  WUE [exposure,pos="-1.000,0.000"]
  C   -> { FL TM }
  CS  -> { IGP PI }
  FL  -> { NMF PGP }
  IGP -> I
  NMF -> { I IGP }
  PGP -> WUE
  TM  -> { IGP PI WUE}
  WUE -> IGP
  }
')

## Plotting amended DAG
plot(g.amended)
```

## 13
{PGP, TM}
{FL, TM}
{PGP, TM, C}
```{r}
adjustmentSets(g.amended, type = 'all')
```

## 14
{PGP, TM}
{FL, TM}
{NMF, TM}
```{r}
adjustmentSets(g.amended, type = 'minimal')
```
# Additional (take-home) exercise
```{r}
## Generate and plot data
set.seed(1234)
n = 1000
x = rnorm(n, 0, 1)
y = x + rnorm(n, 0, 1)
z = y + rnorm(n, 0, 0.1)
DAT = data.frame(x,y,z)
xyplot = ggplot(DAT, aes(x = x, y = y)) + geom_point()  
zyplot = ggplot(DAT, aes(x = z, y = y)) + geom_point()  
grid.arrange(xyplot, zyplot, ncol = 2)
```

## 15
```{r}
lm1 = lm(y ~ x, data = DAT)
summary(lm1)
```
```{r}
lm2 = lm(y ~ z, data = DAT)
summary(lm2)
```
Z would be a better predictor for Y, because the association between Z and Y is stronger.

## 16
A good predictor in the regression sense doesn't always have a good causal effect on the outcome of interest.
